---
title: "Data Exploration"
output: html_notebook
---

# Overview
This document will be used for the data exploration phase of this project.  The first thing that I am going to do is upload the usual code to set up the workspace.

```{r}
# clear R's environment
rm(list = ls())

# upload the libraries that we will need for this analysis
library(ggplot2)
library(dplyr)
library(here)
library(ggfortify)
```
Now I will pull in the dataset that I selected for the capstone project.  The data set consists of different gait analysis metrics for individuals walking up and down a sloped surface with a prosthetic containing a microcontroller and one that does not. 

```{r}
# import data set using read.csv() and stringsAsFactors = TRUE to import the categorical data as factors with levels that we can look at
DF <- read.csv(here("Data", "prosthetic_analysis.csv"), stringsAsFactors = TRUE)

# take an initial look at the structure of the data set
glimpse(DF)
```

# Data Exploration
Now, the three things that I want to analyse within this data set are: 
1. The propulsion of the different types of prosthetic while walking uphill compared to the intact leg.
2. The dampening power of the different types of prostheses compared to the intact leg. 
3. The relationship between stance and overall swing velocity.

After looking at the data initially, there is a lot of information inside of it.  The first thing that we are going to do is simplify things by taking out the data we want and placing it into a new data frame.

```{r}
# select the data that we want from the current data set
DF_refined <- DF %>% select(c("Subject", "Session", "Side", "Slope","Power.Gen...Preswing", "Power.Gen...Midstance", "Power.Gen...Terminal.stance", "Propulsion", "Braking", "Velocity" ))
```
Now that I have the data that I want to take a look at separated from the rest of the data, I will clean it up a bit to make it easier to understand starting with modifying the column titles. 
```{r}
# clean up the column headers
names(DF_refined) <- c("Subject", "Session", "Side", "Slope", "Power_Gen_Pre", "Power_Gen_Mid", "Power_Gen_Terminal", "Propulsion", "Braking", "Velocity")
```



